---
title: ''
output:
  html_document: 
   always_allow_html: true
  word_document: default
editor_options:
  markdown:
    wrap: 72
---

<P>

<BR>

<CENTER><FONT SIZE = 5, COLOR = "#008000"> <B>ALY6015: Intermediate
Analytics</B> <BR><B>Northeastern University</B> <BR>
<FONT SIZE = 4, COLOR = #008000><B> VLADIMIR SHAPIRO </B></FONT>
<BR><BR> Team Project: Draft Report<BR> </FONT></CENTER>

<BR> <FONT SIZE = 4, COLOR = #E74C3C><B> INTRODUCTION </B></FONT>

<p>

In this project, our primary focus is on sports analytics, specifically
in the context of predicting player wages and analyzing the factors that
significantly impact a player's overall rating. By understanding these
aspects, we can gain valuable insights into player valuation and
performance evaluation in the sports industry.

Q1: Predicting a player's wage based on their attributes and personal
data To answer this question, we aim to develop a predictive model that
can estimate a player's wage using their attributes and personal data.
The attributes may include factors such as age, overall rating, market
value, nationality, position, and other relevant information. M1: Our
chosen method for this analysis is Ridge Regression. Ridge Regression is
a suitable technique as it addresses the issue of overfitting by adding
a penalty term to the loss function. By incorporating the attributes
into the model, we can train it to accurately predict a player's wage.
The dataset will be divided into training and testing sets, and
cross-validation techniques will be applied to optimize the model's
performance. Tuning the regularization parameter (alpha) will help
strike a balance between bias and variance, resulting in a more robust
and accurate wage prediction model.

Q2: Analyzing factors that have the most significant impact on a
player's overall rating To answer this question, we aim to identify the
factors that have the most significant influence on a player's overall
rating. This analysis will help us understand the key contributors to a
player's performance evaluation. M1: For this task, we will employ the
Generalized Linear Model (GLM) analysis. GLM allows us to analyze the
relationship between a player's overall rating and various factors,
including age, nationality, position, and specific attributes like pace,
shooting, passing, and more. By examining these factors and their
coefficients, we can determine the relative importance and significance
of each factor in determining a player's overall rating.

Through these analyses, we can provide valuable insights to sports
organizations, agents, and talent scouts, enabling them to make informed
decisions regarding player wages and performance evaluation.

</p>

```{r Libraries, echo=FALSE, include=FALSE }
# Load required libraries
library(ggplot2)
library(dplyr)
library(tidyr)
library(corrplot)
library(readr)
library(tidyverse)
library(knitr)
library(caret)
library(glmnet)
library(kableExtra)
library(ggcorrplot)
library(corrplot)
library(GGally)
install.packages("fastmap")

```

***ANALYSIS***

***Exploratory Data Analysis (EDA)***

```{r warning=FALSE,message=FALSE}
#Import the data
players_22 <- read_csv("C:/Users/Raheb/Desktop/UNI/players_22.csv")
```

```{r}
# Specify the variables we want to keep
selected_vars <- c("wage_eur", "age", "short_name", "value_eur","overall","nationality_name","player_positions","shooting")

# Create a new dataset with only the selected variables
selected_data <- players_22[, selected_vars]

# Remove rows with any missing values
data_complete <- na.omit(selected_data)
```

```{r}
sorted_table <- data_complete %>%
  arrange(desc(wage_eur))
sorted_table %>%
  kbl(caption = "Table1.Summary of the players with their attributes") %>%
  kable_classic(full_width = F, html_font = "Cambria")%>%
     scroll_box(height = "300px")

summary(data_complete)
```

<br><b>Obsversations:</b>

<p>

The table titled "Summary of the players with their attributes" provides
valuable insights into the attributes of various players. The wage
distribution among the players showcases a range from 270,000 EUR to
350,000 EUR, reflecting the diversity in their earnings. Additionally,
the age of the players spans from 26 to 36 years old, indicating a mix
of experienced and relatively younger individuals. Notable names in the
table include K. De Bruyne, K. Benzema, L. Messi, Casemiro, T. Kroos, R.
Sterling, R. Lewandowski, Cristiano Ronaldo, Neymar Jr, and S. Mané,
each contributing their unique skills and abilities to their respective
teams. The market value, represented by the "value_eur" column,
showcases a range from 45,000,000 EUR to 129,000,000 EUR, indicating the
perceived worth of these players in the market. The overall rating
column provides an insight into their overall performance, with ratings
ranging from 88 to 93, highlighting their exceptional capabilities.
Moreover, the players hail from various nationalities, including
Belgium, France, Argentina, Brazil, Germany, England, Poland, and
Portugal, showcasing the global diversity of talent.

The "player_positions" column illustrates the diverse roles and
positions these players occupy on the field. These positions include CM,
CAM, CF, ST, RW, LW, CDM, and CAM, demonstrating their versatility and
adaptability. Finally, the "shooting" attribute values range from 73 to
94, suggesting varying levels of expertise in their shooting abilities..

</p>

```{r}

# Set up a nice plot layout
par(mar = c(5, 4, 4, 6))  # Adjust margin size

# Create a scatter plot of Wage vs. Age
plot(data_complete$age, data_complete$wage_eur, 
     xlab = "Age", ylab = "Wage (EUR)", main = "Figure1. Relationship between age and wage",
     col = "red", pch = 16)
legend("topright", legend = c("Outliers"), pch = 20, col = "red", pt.cex = 1.2, bty = "n")
```

<br><b>Obsersations:</b>

<p>The scatter plot of Wage vs. Age reveals a general trend where player
wages tend to be higher during their late twenties and early thirties.
However, as players enter their mid-thirties, there is a slight decrease
in wages. The plot also includes outliers representing exceptional cases
where players earn significantly higher or lower wages compared to their
age group.</p>

***Goal (Business Question) #1: Predicting a player's wage based on
their attributes and personal data.***

<p>

Motivation: The prediction of player wages is crucial for sports
organizations, agents, and talent scouts when negotiating contracts and
making financial decisions. By accurately estimating a player's wage
based on their attributes and personal data, stakeholders can make
informed decisions and ensure fair compensation.

Research and Development: To tackle this business question, we will
leverage a machine learning approach called Ridge Regression. Ridge
Regression is suitable for predicting a player's wage as it addresses
the issue of overfitting by introducing a penalty term to the loss
function. This regularization technique helps in achieving a more robust
and accurate model.

```{r}

# Select only the numeric variables
numeric_vars <- c("wage_eur", "age", "value_eur", "overall")
numeric_data <- data_complete[, numeric_vars]



# Print the correlation matrix
cor_plot <- ggcorr(numeric_data,
    nbreaks = 6,
    label = TRUE,
    label_size = 3,
    color = "grey50")

cor_plot <- cor_plot + labs(caption = "Figure2.Correlation Matrix Plot")
print(cor_plot)
```

<br>The development process involves the following steps:

</p>

```{r}
   
# Split the dataset into training and testing sets
set.seed(123) # Set a seed for reproducibility
trainIndex <- createDataPartition(data_complete$wage_eur, p = 0.7, list = FALSE)
trainData <- data_complete[trainIndex, ]
testData <- data_complete[-trainIndex, ]

# Perform Ridge Regression
model <- glmnet(x = as.matrix(trainData[, c("age", "overall", "value_eur")]),
                y = trainData$wage_eur,
                alpha = 0, # Ridge Regression
                standardize = TRUE)

# Choose the optimal lambda value using cross-validation
cvModel <- cv.glmnet(x = as.matrix(trainData[, c("age", "overall", "value_eur")]),
                     y = trainData$wage_eur,
                     alpha = 0, # Ridge Regression
                     standardize = TRUE)

# Retrieve the optimal lambda value
optimalLambda <- cvModel$lambda.min

# Train the final model using the optimal lambda value
finalModel <- glmnet(x = as.matrix(trainData[, c("age", "overall", "value_eur")]),
                     y = trainData$wage_eur,
                     alpha = 0, # Ridge Regression
                     standardize = TRUE,
                     lambda = optimalLambda)

# Make predictions on the testing set
predictions <- predict(finalModel, newx = as.matrix(testData[, c("age", "overall", "value_eur")]), s = optimalLambda)

# Model Evaluation
mse <- mean((predictions - testData$wage_eur)^2)
rmse <- sqrt(mse)
r_squared <- cor(predictions, testData$wage_eur)^2



# Create a table for evaluation metrics
evaluation_table <- data.frame(
  Metric = c("Mean Squared Error (MSE)", "Root Mean Squared Error (RMSE)", "R-squared"),
  Value = c(mse, rmse, r_squared)
)
# Format the table using kable
table_formatted <- kable(evaluation_table, align = "c", caption = "Table 2.Model Evaluation Metrics") %>%
  kable_styling(bootstrap_options = c("striped", "hover"))
table_formatted
```

<b>Obsversations</b>

<p>

The output provided represents the evaluation metrics for the Ridge
Regression model used to predict player wages based on their attributes
and personal data.

1.  Mean Squared Error (MSE): `r mse`
    -   MSE is a measure of the average squared difference between the
        predicted and actual wage values. In this case, the MSE value of
        94985789 suggests that, on average, the squared difference
        between the predicted and actual wages is quite large. A lower
        MSE indicates better model performance.
2.  Root Mean Squared Error (RMSE): `r round(rmse,3)`
    -   RMSE is the square root of the MSE and provides a measure of the
        average difference between the predicted and actual wage values.
        In this case, the RMSE value of 9746.065 indicates that, on
        average, the predicted wage values deviate by approximately
        \$9746.065 from the actual wage values. A lower RMSE indicates
        better model performance.
3.  R-squared: `r round(r_squared,3)`
    -   R-squared, also known as the coefficient of determination,
        measures the proportion of the variance in the dependent
        variable (wage) that can be explained by the independent
        variables (attributes and personal data). In this case, the
        R-squared value of 0.7291393 indicates that approximately 72.9%
        of the variance in the player wages can be explained by the
        model. A higher R-squared value suggests better predictive
        power.

Overall, the evaluation metrics provide insights into the performance of
the Ridge Regression model. While the model explains a significant
portion of the wage variance (as indicated by the R-squared value),
there is room for improvement, as evidenced by the relatively high MSE
and RMSE values. Fine-tuning the model or considering additional
variables may help enhance its performance.

</p>

***Goal (Business Question) #2: Analyzing factors that have the most
significant impact on a player's overall rating.*** Q2: Analyzing
factors that have the most significant impact on a player's overall
rating To answer this question, we aim to identify the factors that have
the most significant influence on a player's overall rating. This
analysis will help us understand the key contributors to a player's
performance evaluation. M1: For this task, we will employ the
Generalized Linear Model (GLM) analysis. GLM allows us to analyze the
relationship between a player's overall rating and various factors,
including age, nationality, position, and specific attributes like pace,
shooting, passing, and more. By examining these factors and their
coefficients, we can determine the relative importance and significance
of each factor in determining a player's overall rate

<p>

Motivation: Player's overall rating is the most intuitive display of
player's performance. According to the player's performance score, it
can be clearly concluded that the overall level of each player has
changed and which players have more comprehensive skills. In addition,
the player's performance score helps fans and clubs assess the player's
status and value.

Research and Development: To tackle this business question, we will
leverage feature selection method to determine an optimal model.
Generalized Linear Model (GLM) analysis is suitable for analyze the
relationship between a player's overall rating and various factors as it
determine the extent to which each of these factors contributes to the
player's overall rating. The stepwise selection method is considered to
be used to get the variables effectively.

```{r}
# Select variables
selected_vars2 <- c("overall","age", "skill_moves", "international_reputation", 
                   "pace", "shooting", "passing", 
                   "dribbling", "defending", "physic")

# Subset dataframe based on selected_vars
df_selected2 <- players_22[, selected_vars2]

# Remove rows with NA values
vars_selected2 <- na.omit(df_selected2)


# Print the correlation matrix
cor_plot2 <- ggcorr(vars_selected2,
    nbreaks = 6,
    label = TRUE,
    label_size = 3,
    color = "grey50")

cor_plot21 <- cor_plot2 + labs(caption = "Figure3.Correlation Matrix Plot")
print(cor_plot21)
```

<br>The development process involves the following steps:

</p>

```{r}
# Split the dataset into training and testing sets
set.seed(123) 
trainIndex2 <- createDataPartition(vars_selected2$overall, p = 0.7, list = FALSE)
trainData2 <- vars_selected2[ trainIndex2, ]
testData2 <- vars_selected2[-trainIndex2, ]

# perform generalized linear model 
full_model <- glm(overall ~ age + skill_moves + international_reputation + pace + shooting + passing + dribbling + defending + physic, data = trainData2, family = gaussian())
summary(full_model)

# Perform stepwise regression
step_model <- glm(overall~., data = trainData2) %>% stats::step(direction = "both") 
summary(step_model)

# Predict on the test data using the final model
predictions2 <- predict(step_model, newdata = testData2)


# Model Evaluation
mse2 <- mean((testData2$overall - predictions2)^2)
rmse2 <- sqrt(mse2)
r_squared2 <- cor(predictions2, testData2$overall)^2

# Create a table for evaluation metrics
evaluation_table2 <- data.frame(
  Metric = c("Mean Squared Error (MSE)", "Root Mean Squared Error (RMSE)", "R-squared"),
  Value = c(mse2, rmse2, r_squared2)
)
# Format the table using kable
table_formatted2 <- kable(evaluation_table2, align = "c", caption = "Table 3.Model Evaluation Metrics") %>%
  kable_styling(bootstrap_options = c("striped", "hover"))
table_formatted2
```

<b>Obsversations</b>

<p>

The output provided represents the evaluation metrics for the
generalized linear model used to predict players' overall score based on
their attributes and personal data.

1.  Mean Squared Error (MSE): `r mse2`
    -   The MSE is 9.958295, which implies there is a certain level of
        error in the predictions, although without a benchmark model for
        comparison, it's hard to assess whether this is high or low.
2.  Root Mean Squared Error (RMSE): `r round(rmse2,2)`
    -   The RMSE is 3.16 which means that on average, the prediction of
        the model is about 3.16 scores away from the actual values.
3.  R-squared: `r round(r_squared2,3)`
    -   An R-squared of 0.779 indicate that the model has a relatively
        high explanatory power. An R-squared of 0.779 means that about
        77.9% of the variability in the player's overall score can be
        explained by the features in the model.

Overall, the use of generalized linear model and stepwise model shows
that all selected variables are suitable to predict players' overall
score. According to the R-squared value, the prediction model is
convincible. From the RMSE, the results of the prediction is accurate.
It can be speculated that add additional variables might enhance the
model's performance.

</p>

<BR> <FONT SIZE = 4, COLOR = #E74C3C><B> CONCLUSION </B></FONT>

<p>

We concentrated on sports analytics in this research, particularly on
estimating player salaries and examining the elements that have a major
influence on a player's overall rating. We wanted to give useful
insights into player value and performance evaluation in the sports
market by comprehending these factors.

We used Ridge Regression to answer the first question, which involved
estimating a player's pay based on their characteristics and personal
information. To calculate a player's pay, the algorithm factored in
characteristics including age, overall rating, market worth, country,
and position. With space for improvement in terms of MSE and RMSE, the
assessment metrics showed that the model explained around 72.9% of the
variance in player earnings. The performance of the model may be
improved by adjusting it or by taking new factors into account. Proper
wage forecasting is crucial for sports organizations, agents, and talent
scouts to make informed decisions and ensure fair compensation.

We used the Generalized Linear Model (GLM) analysis for the second
question, which involved examining the variables that affected a
player's total rating. We established the proportionate weighting of
each aspect in creating a player's total rating by looking at elements
including age, nationality, position, and characteristics like pace,
shooting, and passing. The assessment metrics showed that the GLM model
has a strong explanatory power by explaining roughly 77.9% of the
variation in players' total scores. For the purpose of determining a
player's position and value, it is crucial to comprehend the major
factors that influence performance appraisal.

We have successfully answered these inquiries, giving talent scouts,
agents, and sports organizations insightful information. Fair salary
discussions are made possible by accurate wage forecast, and player
assessment and performance analysis are aided by knowledge of the
variables that affect total ratings.

</p>

<BR> <FONT SIZE = 4, COLOR = #E74C3C><B> REFERENCES </B></FONT>

<p>

1.  sonbarsonbar                    2111 silver badge22 bronze badges,
    user12282991user12282991                    4111 silver badge22
    bronze badges, & GanainGanain                    4111 bronze badge.
    (1965, October 1). *Error in length(obj) : Class name too long in
    \"length.\"* Stack Overflow.
    <https://stackoverflow.com/questions/55342788/error-in-lengthobj-class-name-too-long-in-length>

2.  R: Predict method for linear model fits. (n.d.).
    <https://www.math.ucla.edu/~anderson/rw1001/library/base/html/predict.lm.html>

3.  Johnson, D. (2023, May 27). *Correlation in R: Pearson & Spearman
    Correlation Matrix*. Guru99.
    <https://www.guru99.com/r-pearson-spearman-correlation.html>

4.  Singh, D. (2019, November 12). *Deepika Singh*. Pluralsight.
    <https://www.pluralsight.com/guides/linear-lasso-and-ridge-regression-with-r>
